Exploiting Intrinsic Noise for Heterogeneous Cell Control Under Time Delays and Model Uncertainties

### 2023-07-31 paper.tex
Update to captions.

In principle, the master equation in Eq.\ \ref{CME} could contain an uncountably infinite number of states, and therefore the exact solution as well as the null vector in Eq.\ \ref{SSDist} may not be computable exactly.  To address this issue, we first truncate the system at a finite number for each species and then apply a reflecting boundary condition, resulting in a finite dimensional master equation. To assess the time interval over which this truncation is valid, we also solve the Finite State Projection\cite{Munsky:2006} for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.

\brian[We should use the FSP to compute the error of the truncation for a couple cases.  We can talk about how to do this when we meet.]
\michael[ok!]
%Finite state projections are a powerful way to simulate master equation analysis which enables the assessment of simulation accuracy, but these simulations have challenges measuring the steady state of the system since all probability will always enter error at long enough time scales. To enable control of steady state systems, a truncated FSP which reflects probability back into the system enables this calculation.

\subsection{Quantification and optimization of control performance}\label{sec:Quantification}
In the SIMO control of stochastic processes, a single input is applied simultaneously to all systems at once and a control signal that acts beneficially on one system destabilizes other systems. Stochastic SIMO control inputs which seek to control cells one-cell-at-a-time create increased instability and response variability in the other signals not being controlled.
Thus, even in systems which do a good job at controlling a system cannot independently control individual cells to a desired target point without impacting the variance of others, so an effective controller must strike a balance among the desired behaviors of all cells in the system.

To quantify performance success, we define the {\em steady state performance error}, $J$, as the expected steady state Euclidean distance of the process from the specified target state, $\mathbf{T}$:
 \begin{equation}
 J = \mathbb{E}\{|\mathbf{X}-\mathbf{T}|_2\}.
 \end{equation}
The squared score is easily calculated by applying a linear operator on steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
\begin{eqnarray}
J^2&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
&=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i- T_1)^2 + (j- T_2)^2 +\ldots\right],\nonumber  \\
&=\mathbf{C}\mathbf{P}^*,
\label{Euclid} 
\end{eqnarray}
where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target T, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$. 
As a result of this calculation, $J$ is a non-negative scalar that is zero if any only if $\mathbf{P}^*$ is a delta distribution located exactly at the target vector $\mathbf{T}$.

We consider two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that bases its control signal on simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies only on observations from a single cell:
\begin{equation}
\mathbf{u}^{FAC}=u(x_1,x_2),
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_2),
\end{equation}
where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two. Despite their differences in their observation data, both the FAC and the simpler PAC are optimized to achieve the same goal, namely to drive both cells to their respective set points.

Optimization of $\mathbf{u}^{FAC}$ was performed using a gradient descent method to minimize $J$. Since the square root is a monotonically increasing function, minimizing $J^2$ results in the same control as would minimizing $J$ directly. Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ and walking in a small step $du$ in that direction. The calculation of the gradient is given by
\begin{equation}
\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}}
\end{equation}
Where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
\begin{equation}
A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}
\end{equation}
to find $\frac{dP}{d\mathbf{u}^{FAC}}$.

%Optimization of $\mathbf{u}^{PAC}$ was done by restricting $\frac{d(J^2)}{d\mathbf{u}^{FAC}}$ in ambient space to search a low dimensional subspace with a linear lift operator D. Then the subspace search is given by. 
%\begin{equation}
%\frac{d(J^2)}{d\mathbf{u}^{PAC}}=C \frac{dP}{d\mathbf{u}^{PAC}} = C D \frac{dP}{d\mathbf{u}^{FAC}}  
%\end{equation}

%Root mean squared error was calculated by taking the square root of the score.
%\begin{equation}
%RMSE=\sqrt{J}
%\end{equation}

@@

In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection\cite{XXX}, time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, principle orthogonal decompositions \cite{Vo2019}, or other coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}. 
Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration. 
To explore how mismatches in assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model. 
To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
\begin{equation}\label{eq:granscale}
w'(\mathbf{X})=w(\mathbf{X}/\alpha).
\end{equation}
For example, the production and degradation of protein in cell one would become:
\begin{eqnarray}
w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0 \nonumber \\
 w'_2(\mathbf{X}) &= \gamma x_1/\alpha. \label{eq:rescaling}
\end{eqnarray}
We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the  control level assigned to the current state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$. Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, control signal values at fractional state values after rescaling $(x_1/\alpha,x_2/\alpha)$ are calculated using 2D cubic interpolation from the control values at the nearest integer state values.
%\begin{equation}
%u'=u(x_0 / \alpha, x_1 / \alpha) 
%\end{equation}
%\begin{equation}
%u'=u(\alpha x_0)
%\end{equation}
Finally, to provide a consistent metric for relative scoring, the definition of the performance score is also adjusted according to scale magnitudes. For example, in the two cell system the new steady state performance error would become:
\begin{eqnarray}
(J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*.
\label{EuclidV}
\end{eqnarray}
We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and these are not recomputed or refit upon changing the system granularity.

@@

Delays are inherent in any realistic control systems, and it this case delays would be expected to arise due to the time needed for various biochemical reactions such as formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein \cite{Cai2007}. 
Additional delays would also arise from data analysis, decision making, and actuation dynamics. 
To investigate the effects of time delay on control actuation, we devised a simple time-delay stochastic simulation algorithm. 
This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and the time delayed control input propensity.  
Using this information, the time-delayed control signal at time $t$ can be specified as:

\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\textrm{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\textrm{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$), and $x$ gathered from the state history. 
\brian[Explain how you modify the SSA for the delayed reaction SSA.  Look through the literature to cite someone else who has implenented a similar method.]
We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.

%\subsection{Parameter Scalings and Sensitivity}
%Finally, to explore the sensitivity of the control performance to inaccuracies in model parameters for one or more cells, we calculate the control performance score ($J$), over a range for each parameter scaled linearly with a scaling parameter ($\zeta$) while holding all other parameters constant. 
%%Any new arbitrary parameter $\lambda\in \{ \kappa,\beta,k_0,\eta,\gamma\}$ can be rescaled from an old one ($\lambda$) according to
%%\begin{equation}
%%\lambda' = s \lambda
%%\end{equation}
%We explore two types of sensitivity analysis: one which performs the analysis on one cell at a time while keeping the other constant, and another which scales similar parameters in both cells at the same time. 
%%In the former case, it should be noted that each cell receives its own set of parameters, so that the number of parameters double during sensitivity analysis of a two cell control process. 
%In either case, steady state distributions of the perturbed models are solved using the FSP.
% XXX - I don't think that we need this section in Methods -- but perhaps some of the text should be moved to the results if needed.

\subsection{Dynamic Setpoint Control}\label{sec:Dynamic}
Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq. \ref{CME}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study. 
To circumvent this challenge, we instead propose a simple alternative to sweep though a set of controllers over time which target different static points in piecewise fashion to mimic a dynamically moving path.

 We start by describing a discrete path  of target points in state space which cycle at fixed frequency.
  Then, we optimize controllers for each target point and cycling through the controllers over time to create the final controller $\mathbf{R}(\mathbf{x},t)$.
Let $\mathbf{p}(t)$ be a continuous and cyclicly moving point in state space
\begin{equation}
\mathbf{p(t)}=[x_1(t),x_2(t)]
\end{equation}
where $x_1(t)$, and $x_2(t)$ are both cyclic functions of time $\in \mathbf{R}$.
Then, $\mathbf{q}$ is a collection of points $q_i$ which evaluate the cyclic function over its natural period (T) in n steps. 
\begin{equation}
\mathbf{q}=[q_1,\ldots ,q_n]=[\mathbf{p}(t_1),\ldots,\mathbf{p}(t_n)]
\end{equation}
where each $t_i$ comes from linearly spaced points between $0$ and $T$, in $n$ steps 
\begin{equation}
t_i = [t_1,\ldots,t_n]=[0,\frac{1}{n} T,\ldots,\frac{n}{n}T]
\end{equation}
Then, we optimize a set of controllers which target each $q_i$ in the discretized path.
\begin{equation}
\mathbf{U}_i(\mathbf{x})=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
\end{equation}
Next, we cycle through each controller at a fixed frequency by using ($\phi(t)$). 
\begin{equation}
\phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
\end{equation}
where $f$ the cycling frequency given by $1/T$. Then the final control equation is given by 
 \begin{equation}
\mathbf{R}(\mathbf{x},t)=\bf{U}_{\phi(t)}(x)
\end{equation}
Where $\mathbf{R}(\mathbf{x},t)$ returns the control input scalar as a function of state and time. 
The dynamical system equation becomes
\begin{equation}
\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\mathbf{R}(\mathbf{x},t)
\end{equation}
Since $\mathbf{R}(\mathbf{x},t)$ is a function of time, the system is no longer time invariant. 
\section{Results}

\brian[{\bf We still need this.} Add a paragraph or two to explain the big picture here.  Use figure 1 to remind the reader of (1) the control goals, (2) the control law, (3) the resulting distributions, (4) the objective function and (5) the optimization of the control law to minimize the error.  Refer to the methods for each such calculation.  This section is critical to make sure the reader understands what is coming next.  Then provide an outline of the remaining subsections of the results section so that the reader knows what to expect.]

\michael[new paragraph here]

In this section, we seek to show that noise-exploiting controllers are effective at controlling a noisy system to a dynamically moving target point. To do this, we show that control performance is robust to a range of parameter uncertainties without the needing adjustments to the controller. Then, we extend the control performance analysis to understand how time delay and system scale alters control performance. Furthermore, we find regions in target space which are easiest to control by optimizing individual controllers for each target. We extend these analysis to change the target point over time in a Peicewise fashion. 
 \subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}
\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). Columns show each parameter in the model, rows show the cell which has its parameter perturbed. }
\label{Parameter}
\end{center}
\end{figure*}

Real stochastic processes always have unknown or uncertain mechanisms and parameter values. Although model structures and parameter estimates can often be obtained through fitting to training data, these estimates will never be perfect due to unavoidable measurement errors or limited data sets. 
Even with plentiful and precise training data, model and parameter uncertainties are inevitable due to the fact that even genetically identical single-cells exhibit heterogeneity in parameters due to extrinsic variations.
To asses the control performance under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
For each parameter, $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$(dashed lines) control performance was examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Figure.\ \ref{Parameter}B shows the control performance as single parameters in Cell 2 is varied. 
Finally, \ref{Parameter}C shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time.

Modification of parameters were found to produce a broad range of effects. 
For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Figure.\ \ref{Parameter}A and \ref{Parameter}B). 
In some cases, the effects on performance are not monotonic; for example, increasing $\kappa$ in Cell 1 (Figure.\ \ref{Parameter}A, leftmost column) would be highly advantageous up to a limit after which the control performance degrades rapidly. 
In other cases (such as for the promoter leakage rate, $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=10) perturbations, suggesting that this parameter is not important to control performance.
Figure \ref{Parameter}C shows that even when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance. 
In particular, the analysis shows that control performance could be improved by modifying the system design either to increase the auto-regulation promoter strength ($\kappa$) or its cooperativity ($\eta$) or to decrease the promoter binding constant ($\beta$) or the protein degradation rate ($\gamma$).

\subsection{Controllers trained using one assumed level of granularity can remain effective at other levels of granularity.}

We next explored how the granularity of the system would affect the differential control of the process. Specifically, we define the level of granularity using the parameter $\alpha$ as described in Section \ref{sec:Scaling} and which affects the propensity functions and scoring according to equations \ref{eq:granscale}-\ref{EuclidV}. Effectively, larger $\alpha$ corresponds to systems where larger numbers of individual molecules are needed to achieve the same concentration (e.g., larger volumes), while smaller $\alpha$ corresponds to smaller population sizes (e.g., smaller volumes). We previously optimized the controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ based on a the default assumption that $\alpha=1$, and we wished to know what would be the consequences if this same controller were to be applied to a system that has a different granularity.

>>>>

In principle, the master equation in Eq.\ \ref{CME} could contain an uncountably infinite number of states, and therefore the exact solution as well as the null vector in Eq.\ \ref{SSDist} may not be computable exactly.  To address this issue, we first truncate the system at a finite number for each species and then apply a reflecting boundary condition, resulting in a finite dimensional master equation. To assess the time interval over which this truncation is valid, we also solve the Finite State Projection \brian[\cite{Munsky:2006}] for the same truncation, which allows us to compute an upper bound on the truncation error as a function of time.

\brian[We should use the FSP to compute the error of the truncation for a couple cases.  We can talk about how to do this when we meet.]
\michael[ok!]

\subsection{Quantification and optimization of control performance}\label{sec:Quantification}
In the SIMO control of stochastic processes, a single input is applied simultaneously to all systems at once, and therefore a control signal that acts beneficially on one cell may destabilize other cells. 
An effective controller must strike a balance among the desired behaviors of all cells in the system. 
To quantify overall performance success, we define the {\em steady state performance error}, $J$, as the expected steady state Euclidean distance of the process from the specified target state, $\mathbf{T}$:
 \begin{equation}
 J = \mathbb{E}\{|\mathbf{X}-\mathbf{T}|_2\}.
 \end{equation}
The squared score is easily calculated by applying a linear operator to the steady state probability distribution $\mathbf{P}^*$ (Eq.\ \ref{SSDist}) as follows:
\begin{eqnarray}
J^2&= \lim_{t\rightarrow \infty}\mathbb{E}\{|\mathbf{X}(t)-\mathbf{T}|_2^2\}, \nonumber \\ 
&=\sum_{i_1,i_2,\ldots} P^*(x_1=i_1,x_2=i_2,\ldots) \left[(i- T_1)^2 + (j- T_2)^2 +\ldots\right],\nonumber  \\
&=\mathbf{C}\mathbf{P}^*,
\label{Euclid} 
\end{eqnarray}
where $\mathbf{C}$ is simply a vector that contains the squared Euclidian distance of each state from the specified target $\mathbf{T}$, i.e., $C_i = |\mathbf{X}_i-\mathbf{T}|_2^2$. 
As a result of this calculation, $J$ is a non-negative scalar that is zero only if $\mathbf{P}^*$ is a delta distribution located exactly at the target vector $\mathbf{T}$.

We consider two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that bases its control signal on simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies only on observations from a single cell:
\begin{equation}
\mathbf{u}^{FAC}=u(x_1,x_2),
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_2),
\end{equation}
where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the instantaneous number of proteins in cell one and cell two. Despite their differences in their observation data, both the FAC and the simpler PAC are optimized to achieve the same goal, namely to drive both cells to their respective set points.

Optimization of $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were performed using a gradient descent method to minimize $J$. Since the square root is a monotonically increasing function, minimizing $J^2$ results in the same control as would minimizing $J$ directly. Therefore, we calculate the negative gradient $-\frac{d(J^2)}{d\mathbf{u}^{(.)}}$ and adjust parameters a small step $d\mathbf{u}^{(.)}$ in that direction. For example, the calculation of the gradient for the FAC controller is given by:
\begin{equation}
\frac{d(J^2)}{d\mathbf{u}^{FAC}}=C \frac{dP}{d\mathbf{u}^{FAC}},
\end{equation}
where $\frac{dP}{d\mathbf{u}^{FAC}}$ can be solved using general minimized residual calculation of
\begin{equation}
A \frac{dP}{d\mathbf{u}^{FAC}}=B \mathbf{u}^{FAC}.
\end{equation}

@@

In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection \brian[\cite{XXX}], time scale separations \brian[\cite{Peles2006}], Krylov subspaces \brian[\cite{Macnemara&Sidje?}], principle orthogonal decompositions \brian[\cite{Vo2019}], or other coarse meshes \brian[\cite{Munsky2008IEEE,Tapia2012CDC}]. 
Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration. 
To explore how mismatches in the assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model. 
To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
\begin{equation}\label{eq:granscale}
w'(\mathbf{X})=w(\mathbf{X}/\alpha).
\end{equation}
For example, the production and degradation of protein in cell one would become:
\begin{eqnarray}
w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0, \nonumber \\
 w'_2(\mathbf{X}) &= \gamma x_1/\alpha. \label{eq:rescaling}
\end{eqnarray}
We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the  control level assigned to the current state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$. Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, control signal values at fractional state values after rescaling $(x_1/\alpha,x_2/\alpha)$ are calculated using 2D cubic interpolation from the control values at the nearest integer state values.
Finally, to provide a consistent metric for relative scoring, the definition of the performance score is also adjusted according to scale magnitudes. For example, in the two cell system the new steady state performance error would become:
\begin{eqnarray}
(J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*.
\label{EuclidV}
\end{eqnarray}
We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and to simulate a practical application where scales may be unknown or variable, these are not recomputed or refit upon changing the system granularity.

@@

Delays are inherent to any realistic control system, and in this case delays would be expected to arise due to the time needed for various biochemical reactions such as the formation of complete polymerases, activation of promoters, transcription and transport of mature mRNA, and the translation and maturation of protein \cite{Cai2007}. 
Additional delays would also arise from data analysis, decision making, and actuation dynamics. 
To investigate the effects of observation or actuation time delays on control performance, we devised a simple time-delay stochastic simulation algorithm. 
This algorithm records the state history after each stochastic reaction, enabling reconstruction of the population history of the species and the time delayed control input propensity.  
Using this information, the time-delayed control signal at time $t$ can be specified as:

\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\textrm{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\textrm{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$), and $x$ gathered from the state history. 
We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.
\brian[Still missing reference to the extrande method at this website
\url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004923}].

\subsection{Tracking Time-Varying Trajectories}\label{sec:Dynamic}
Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq. \ref{CME}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study. 
To circumvent this challenge, we instead propose a simple alternative in which the controller sweep though a piecewise constant set of controllers each designed for a specific static target point along the desired trajectory.

Our goal is to control the system to follow a specific target trajectory, $\mathbf{T}(t)$. We choose a discrete set of $K$ target points along this trajectory:
\begin{equation}
\left\{\mathbf{T}_1,\mathbf{T}_2,\ldots,\mathbf{T}_K\right\},
\end{equation}
and for each individual target $\mathbf{T}_i$, we optimize to find a corresponding controller, $\mathbf{u}_i=\mathbf{u}(\mathbf{T}_i)$. Finally, to implement the control at any given time, $t$, we find the index of the nearest precomputed target state:
\begin{equation}
\hat{i}(t) = \textrm{argmin}_i \left| \mathbf{T}(t) -  \mathbf{T}_i\right|_2,
\end{equation}
and assign that controller.
%
%controllers for each target point and cycling through the controllers over time to create the final controller $\mathbf{R}(\mathbf{x},t)$.
%Let $\mathbf{p}(t)$ be a continuous and cyclicly moving point in state space
%\begin{equation}
%\mathbf{p(t)}=[x_1(t),x_2(t)]
%\end{equation}
%where $x_1(t)$, and $x_2(t)$ are both cyclic functions of time $\in \mathbf{R}$.
%Then, $\mathbf{q}$ is a collection of points $q_i$ which evaluate the cyclic function over its natural period (T) in n steps. 
%\begin{equation}
%\mathbf{q}=[q_1,\ldots ,q_n]=[\mathbf{p}(t_1),\ldots,\mathbf{p}(t_n)]
%\end{equation}
%where each $t_i$ comes from linearly spaced points between $0$ and $T$, in $n$ steps 
%\begin{equation}
%t_i = [t_1,\ldots,t_n]=[0,\frac{1}{n} T,\ldots,\frac{n}{n}T]
%\end{equation}
%Then, we optimize a set of controllers which target each $q_i$ in the discretized path.
%\begin{equation}
%\mathbf{U}_i(\mathbf{x})=[\mathbf{u}_1(\mathbf{x}),...,\mathbf{u}_n(\mathbf{x})]
%\end{equation}
%Next, we cycle through each controller at a fixed frequency by using ($\phi(t)$). 
%\begin{equation}
%\phi(t)=\lceil \frac{n}{T} mod(t,T)\rceil = \lceil n \ f \ mod(t,\frac{1}{f})\rceil
%\end{equation}
%where $f$ the cycling frequency given by $1/T$. 
%Then the final control equation is given by 
% \begin{equation}
%\mathbf{R}(\mathbf{x},t)=\bf{U}_{\phi(t)}(x)
%\end{equation}
%Where $\mathbf{R}(\mathbf{x},t)$ returns the control input scalar as a function of state and time. 
Under this piecewise constant controller law, the full time-varying CME becomes
\begin{equation}
\frac{d}{dt}\mathbf{P}=\mathbf{A}+\mathbf{B}\mathbf{u}_{\hat{i}(t)}(\mathbf{x}).
\end{equation}
\section{Results}

%\brian[{\bf We still need this.} Add a paragraph or two to explain the big picture here.  Use figure 1 to remind the reader of (1) the control goals, (2) the control law, (3) the resulting distributions, (4) the objective function and (5) the optimization of the control law to minimize the error.  Refer to the methods for each such calculation.  This section is critical to make sure the reader understands what is coming next.  Then provide an outline of the remaining subsections of the results section so that the reader knows what to expect.]

As developed in the Methods section, we focus on a SIMO controller where {\em all cells receive the same input} at every instant of time. For this, the control signal depends upon the observed state, e.g., $u(t) = \mathbf{u}(\mathbf{x})$ as developed above and illustrated in Fig. \ref{cartoons}(I,III). These controllers have been optimized to break symmetry so that multiple cells can be controlled to different targets as illustrated in Fig. \ref{cartoons}(II,IV), and we quantify performance according to the RMSE error introduced in Eq. \ref{Euclid}. To explore how such controllers may perform in realistic settings where the models are approximate and parameter are unknown or extrinsically variable, we now fix the parameters of those controllers and explore performance robustness to different types of model uncertainties, including parameter errors or variations (Section \ref{sec:ResPars}), incorrect assumptions on system scales (Section \ref{sec:ResGran}) and time delays (Section \ref{sec:ResTime}). Finally, in Section \ref{sec:ResTraj}, we extend the control analysis to consider the performance of the controller for tracking variable reference signals with different frequencies and phases.

 \subsection{Stochastic SIMO Optogenetic Control can Remain Effective Despite Small Parameter Errors or Extrinsic Uncertainties}\label{sec:ResPars}
\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). Columns show each parameter in the model, rows show the cell which has its parameter perturbed. }
\label{Parameter}
\end{center}
\end{figure*}

Real stochastic processes always have unknown or uncertain mechanisms and parameter values. Although model structures and parameter estimates can often be obtained through fitting to training data, these estimates will never be perfect due to unavoidable measurement errors or limited data sets. 
Even with plentiful and precise training data, model and parameter uncertainties are inevitable due to the fact that even genetically identical single-cells exhibit heterogeneity in parameters due to extrinsic variations.
To asses the control performance under such parameter errors, we performed sensitivity analysis on each model parameter for each cell, then on both cells simultaneously. 
For each parameter, $\mathbf{u}^{\rm FAC}$ (solid lines) and $\mathbf{u}^{\rm PAC}$(dashed lines) control performance was examined across a parameter perturbation range from one-tenth to ten-fold of the original values.
Figure \ref{Parameter}A shows the results for these parameter sweeps when a single parameter in Cell 1 is varied (all other parameters are fixed), and Figure.\ \ref{Parameter}B shows the control performance when a single parameter in Cell 2 is varied. 
Finally, \ref{Parameter}C shows the performance change when a single parameter is changed simultaneously in both Cell 1 and Cell 2 at the same time.

Modifications of parameters were found to produce a broad range of effects. 
For example, increasing $\beta$ in Cell 1 quickly worsens performance while increasing $\beta$ in Cell 2 improves performance (compare second column in Figure.\ \ref{Parameter}A and \ref{Parameter}B). 
In some cases, the effects on performance are not monotonic; for example, increasing $\kappa$ in Cell 1 (Figure.\ \ref{Parameter}A, leftmost column) would be highly advantageous up to a limit after which the control performance degrades rapidly. 
In other cases (such as for the promoter leakage rate, $k_0$), the effect of parameter perturbations on performance is insignificant even for relatively large ($s$=10) perturbations, suggesting that this parameter is not important to control performance.
Figure \ref{Parameter}C shows that even when parameters of both Cell 1 and Cell 2 are jointly changed, these changes could also improve or detract from control performance. 
In particular, the analysis shows that control performance could be improved by modifying the system design either to increase the auto-regulation promoter strength ($\kappa$) or its cooperativity ($\eta$) or to decrease the promoter binding constant ($\beta$) or the protein degradation rate ($\gamma$). We note that co-optimization of both the system and the controller law would allow for further improvements to the performance.

\subsection{Controllers trained using one assumed level of granularity can remain effective at other levels of granularity.}\label{sec:ResGran}

We next explored how the granularity of the system would affect the differential control of the multi-cell system. Specifically, we define the level of granularity using the parameter $\alpha$ as described in Section \ref{sec:Scaling} and which affects the propensity functions and scoring according to Eqns.\ \ref{eq:granscale}-\ref{EuclidV}. Effectively, larger $\alpha$ corresponds to systems where larger numbers of individual molecules are needed to achieve the same concentration (e.g., larger volumes), while smaller $\alpha$ corresponds to situations where smaller population sizes can achieve that concentration (e.g., smaller volumes). We previously optimized the controllers $\mathbf{u}^{\rm FAC}$ and $\mathbf{u}^{\rm PAC}$ based on a the default assumption that $\alpha=1$, and we wished to know what would be the consequences if this same controller were to be applied to a system that has a different granularity.
